Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-06 implementation.tex
Addressed a bunch of Drew's margin comments
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@945 ccdd05d5-56c1-461a-bf98-85f6876523b4

The \fir\ filter banks are implemented using a \gstlal\ element called {\tt
lal\_firbank}, which produces $N$ channels of filter output from an $N\times M$ matrix of
\fir\ filter coefficients.  This element is used in parallel branches in the
pipeline to implement the \SVD\ basis filters in each time
slice.  Rather than implement the time-sliced templates as zero-padded \fir\
filters as described in \eqref{eq:time-slices} we instead implement them as shorter
filters that contain only the nonzero samples.  Adding the appropriate time
offset to the filter output later in the pipeline makes up for the lack of
explicit zero-padding.
%
\editorial{Drew: This final sentence should be removed. It belongs to the next section.}
%
The orthogonal filter outputs must next be reconstructed
into the output of the underlying time-sliced templates.

>>>>

The \fir\ filtering is implemented using a \gstlal\ element called {\tt
lal\_firbank}, which produces $N$ channels of filter output from an $N\times M$ matrix of
\fir\ filter coefficients, representing the \SVD{} basis filters in our application.  This element is used in parallel branches in the
pipeline to implement the \SVD\ basis filters in each time
slice.  Rather than implement the time-sliced templates as zero-padded \fir\
filters as described in \eqref{eq:time-slices} we instead implement them as shorter
filters that contain only the nonzero samples.  Adding the appropriate time
offset to the filter output later in the pipeline makes up for the lack of
explicit zero-padding.

### 2011-06-06 method.tex
Addressed a bunch of Drew's margin comments
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@945 ccdd05d5-56c1-461a-bf98-85f6876523b4

In the remainder of this section we explore a method for reducing the
computational cost of a \TD\ search for compact binary coalescence.  We will
give a truly zero-latency algorithm that competes in terms of floating point
operations per second with the conventional \FD\ method, which requires a
signficant latency in order to be computationally cheap. Our method, \lloid{},
involves two transformations of the template waveforms that produce a set of
orthogonal filters with far fewer coefficients than the original templates.
The reduction in sample count for all of the filters required to search the
entire parameter space is dramatic.
%
\editorial{Drew: is ``sample count" the correct wording?}

The first transformation is to chop the time-domain templates up into time
slices.  Since each template slice is disjoint in time, the resulting set for a
single template is orthogonal.  Given the chirp-like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling removes a factor of $\sim$100 samples
from
%
\editorial{Where do these numbers come from? CHAD: I have actually changed the statement to 100 for the early time slices.}
%
the early part of the waveform and allows the filters to be evaluated at about
$\sim$100 times lower rate.  This amounts to more than a factor of 10000
reduction in the floating-point operations per second required to filter the
early part of the waveform.  However, the resulting filters are still not
orthogonal across the mass parameter space, and are in fact highly redundant.
We use the singular value decomposition to produce an orthogonal filter set
from the time-sliced templates~\cite{Cannon:2010p10398}.  We find that this
reduces the number of sample points of the filters by another factor of $\sim$100.
The combined methods reduce the number of floating point operations to
the level where they are competitive with the conventional high-latency \FD\
matched filter approach.  In the remainder of this section 
%
\editorial{Drew: We used this wording at the beginning of this paragraph.}
%
we describe the
\lloid\ algorithm in detail and provide some basic computational cost scaling.

@@

The first step of our proposed method is to divide the templates into
\emph{time slices} 
%
\editorial{Drew: This was introduced in the previous section. It should be italicized there.}
%
in a time-domain analogue to the frequency-domain
decomposition described in~\cite{Marion2004, Buskulic2010}.
\editorial{nvf: I've picked the two most relevant; I think they are reasonably complementary.}%
A matched filter is constructed for each time
slice.  The outputs form an ensemble of partial \SNR{} streams.  By linearity,
these partial \SNR\ streams can be suitably time-delayed and summed to
reproduce the \SNR\ of the full template.  To wit,
%
\begin{equation}
\label{eq:time-slices}
h_{i}[k] = \sum_{s=0}^{S-1}
	\cases{%
		h_i^s[k] & for $t^s \leqslant k / f^0 < t^{s+1}$ \\
		0 & otherwise
	}
\end{equation}
\editorial{Drew: Should this be $h_{i}[k]$ or $\rho_{i}[k]$?}
%
for $S$ integers $\{f^0 t^s\}$ such that $0  = f^0 t^0 < f^0 t^1 < \cdots < f^0 t^S = N$.
We will show
in the next section that this, combined with the singular value docomposition,
is sufficient to enable a computationally efficient time-domain search and
furthermore is an essential part of an \earlywarning\ detection scheme.

@@

We now examine the computational cost scaling of the approximate implementation
of a conventional time-domain or frequency-domain matched filter bank as compared
with \lloid{}.  For convenience, table~\ref{tab:recap} provides a review of the notation
that we will need in this section.
%
% symbols used in FLOPs calculations table
%
\Table{\label{tab:recap} Notation used to describe filters.}
\br
	& Definition \\
\mr
\numtmps		& number of templates \\
\tmpsamps	& number of samples per template \\
\numslices	& number of time slices \\
\numsvdtmps	& number of orthogonal templates in time slice $s$ \\
\slicessamps	& number of samples in time slice $s$ \\
$f^s$		& sample rate in time slice $s$ \\
$N^\shortdownarrow$ & number of coefficients in decimation filter \\
$N^\shortuparrow$ & number of coefficients in interpolation filter \\
\br
\end{tabular}
\end{indented}
\end{table}
\editorial{Drew: Is \slicessamps before or after decimation? We should clarify that here.}

@@

\Table{\label{table:flops}Computational cost in \flops\ of the \TD\ method, the \FD\ method, and \lloid\ for the bank of $2 \times 657$ templates sampled at 4096 Hz for chirp masses $1.955 \leqslant \mathcal{M} / M_\odot \leqslant 1.2045$.  A block size of $D = 2 N$ is used for the \fft{}s in the \FD\ method.  Both the \FD\ method and \lloid\ are five orders of magnitude faster than the conventional \TD\ method.  However, the \FD\ method has a latency of over half of an hour, whereas \lloid\ has no latency at all.}
\br
Method & \flops\ & latency (s) \\
\mr
Time domain (\TD) convolution & $2.4\times10^{13}$ & 0 \\
Frequency domain (\FD) convolution & $2.6\times10^8$ & 2201 \\
\lloid\ (this work) & $4.7\times10^8$ & 0 \\
\br
\end{tabular}
\end{indented}
\end{table}
\editorial{Drew:``$1.955 \leqslant \mathcal{M} / M_\odot \leqslant 1.2045$" can't be true. Should the signs be greater than signs?}

>>>>

Here we explore a method for reducing the
computational cost of a \TD\ search for compact binary coalescence.  We will
give a truly zero-latency algorithm that competes in terms of floating point
operations per second with the conventional \FD\ method, which requires a
signficant latency in order to be computationally cheap. Our method, \lloid{},
involves two transformations of the template waveforms that produce a set of
orthogonal filters with far fewer coefficients than the original templates.

The first transformation is to chop the time-domain templates up into \emph{time
slices}.  Since each template slice is disjoint in time, the resulting set for a
single template is orthogonal.  Given the chirp-like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling removes a factor of $\sim$100 samples
from
%
\editorial{Where do these numbers come from? CHAD: I have actually changed the statement to 100 for the early time slices.}
%
the early part of the waveform and allows the filters to be evaluated at about
$\sim$100 times lower rate.  This amounts to more than a factor of 10000
reduction in the floating-point operations per second required to filter the
early part of the waveform.  However, the resulting filters are still not
orthogonal across the mass parameter space, and are in fact highly redundant.
We use the singular value decomposition to produce an orthogonal filter set
from the time-sliced templates~\cite{Cannon:2010p10398}.  We find that this
reduces the number of sample points of the filters by another factor of $\sim$100.
The combined methods reduce the number of floating point operations to
the level where they are competitive with the conventional high-latency \FD\
matched filter approach.  In the remainder of this section we describe the
\lloid\ algorithm in detail and provide some basic computational cost scaling.

@@

The first step of our proposed method is to divide the templates into
time slices in a time-domain analogue to the frequency-domain
decomposition described in~\cite{Marion2004, Buskulic2010}.
A matched filter is constructed for each time
slice.  The outputs form an ensemble of partial \SNR{} streams.  By linearity,
these partial \SNR\ streams can be suitably time-delayed and summed to
reproduce the \SNR\ of the full template.  To wit,
%
\begin{equation}
\label{eq:time-slices}
h_{i}[k] = \sum_{s=0}^{S-1}
	\cases{%
		h_i^s[k] & for $t^s \leqslant k / f^0 < t^{s+1}$ \\
		0 & otherwise
	}
\end{equation}
\editorial{Drew: Should this be $h_{i}[k]$ or $\rho_{i}[k]$?}
%
for $S$ integers $\{f^0 t^s\}$ such that $0  = f^0 t^0 < f^0 t^1 < \cdots < f^0 t^S = N$.
We will show
in the next section that this, combined with the singular value docomposition,
is sufficient to enable a computationally efficient time-domain search and
furthermore is an essential part of an \earlywarning\ detection scheme.

@@

We now examine the computational cost scaling of the approximate implementation
of a conventional time-domain or frequency-domain matched filter bank as compared
with \lloid{}.  For convenience, table~\ref{tab:recap} provides a review of the notation
that we will need in this section.
%
% symbols used in FLOPs calculations table
%
\Table{\label{tab:recap} Notation used to describe filters.}
\br
	& Definition \\
\mr
\numtmps		& number of templates \\
\tmpsamps	& number of samples per template \\
\numslices	& number of time slices \\
\numsvdtmps	& number of orthogonal templates in time slice $s$ \\
\slicessamps	& number of samples in decimated time slice $s$\\
$f^s$		& sample rate in time slice $s$ \\
$N^\shortdownarrow$ & number of coefficients in decimation filter \\
$N^\shortuparrow$ & number of coefficients in interpolation filter \\
\br
\end{tabular}
\end{indented}
\end{table}

@@

\Table{\label{table:flops}Computational cost in \flops\ of the \TD\ method, the \FD\ method, and \lloid\ for the bank of $2 \times 657$ templates sampled at 4096 Hz for chirp masses $1.1955 \leqslant \mathcal{M} / M_\odot \leqslant 1.2045$.  A block size of $D = 2 N$ is used for the \fft{}s in the \FD\ method.  Both the \FD\ method and \lloid\ are five orders of magnitude faster than the conventional \TD\ method.  However, the \FD\ method has a latency of over half of an hour, whereas \lloid\ has no latency at all.}
\br
Method & \flops\ & latency (s) \\
\mr
Time domain (\TD) convolution & $2.4\times10^{13}$ & 0 \\
Frequency domain (\FD) convolution & $2.6\times10^8$ & 2201 \\
\lloid\ (this work) & $4.7\times10^8$ & 0 \\
\br
\end{tabular}
\end{indented}
\end{table}
